---
title: "Sampling"
---

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```



### Task 2: Load 1,000 rows of data with the ages of people from 18-80

```{r}
set.seed(42)

population <- tibble(
  age = sample(18:80,1000, replace = TRUE)
)
population
```



### Task 3: Sample 100 rows of data from the dataset

```{r}
sampled_data <- population |> slice_sample(n=15)
sampled_data
```


### Task 4: Calculate the mean age and standard deviation of the sample

```{r}
sample_stats <- sampled_data |>
  summarise(
    mean_age = mean(age),
    sd_age = sd(age)
  )
sample_stats
```



### Task 5: Calculate the mean age and standard deviation of the population

```{r}
population_stats <- population |>
  summarise(
    mean_age = mean(age),
    sd_age = sd(age)
  )
population_stats
```



### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

When comparing the means and standard deviation, they look relativley similar and arre around 1-2 points off from each other. When I changed the sample size from 30 to 15 I didnt see a result change in the standard deviation, except for the mean of the sampled data which went down about 4 points.
