---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Sol Garcia-Terrazas"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
health_inspectionsc <- read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
View(health_inspectionsc)

```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- health_inspectionsc |> 
  summarize(mean_score = mean(compliance_score),
            median_score = median(compliance_score),
            min_score = min(compliance_score),
            max_score = max(compliance_score))
print(compliance_stats)
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
health_inspectionsc |>
  ggplot() +
  geom_histogram(aes(x = compliance_score), binwidth = 4) +
  geom_vline(aes(xintercept = mean(compliance_score, na.rm = TRUE)),
             color = "red", linetype = "dashed", size =1)+
  geom_vline(aes(xintercept = mean(compliance_score, na.rm = TRUE) - sd(compliance_score, na.rm = TRUE)),
             color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(compliance_score, na.rm = TRUE) + sd(compliance_score, na.rm = TRUE)),
             color = "blue", linetype = "dashed", size = 1) +
            labs(
    title = "Histogram of Restaurant Compliance Scores",
    x = "Compliance Score",
    y = "Number of Inspections"
  ) +
  theme_minimal()
```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**Answer:**
The histogram shows that most restaurant compliance score are high, with a large concentration on the 100 end it creates a left-skewed distribution. The standard deviation shows that most restaurant score near the average of 96, although there is variation in the small number of restaurants receiving very low compliance scores. My newsworthy angle would be to focus on the trends of low scoring restaurants, what violations are most frequent, can we see spot a pattern among these low scoring restaurants.
---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
wmata_ridership <- read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
View(wmata_ridership)
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- wmata_ridership |>
            
 summarise(mean_bus <- mean(bus),
            sd_bus <- sd(bus),
            mean_rail <- mean(rail),
            sd_rail <- sd(rail))

print(wmata_stats)
```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- wmata_ridership |>
  mutate(
  rail_zscore = (rail- mean(rail)) / sd(rail) , 
  bus_zscore =  (bus - mean(bus)) / sd(bus)
  )
print(wmata_zscores)

```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
unusual_days <- wmata_zscores |>
  filter(abs(rail_zscore)>2) |>
  arrange(desc(abs(rail_zscore)))
    
print (unusual_days)
```

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(42)

sample_91 <- wmata_ridership |> sample_n(91)

sample_stats <- sample_91 |>
  summarise(
    mean_bus <- mean(sample_91$bus),
    sd_bus <- sd(sample_91$bus),
    mean_rail <- mean(sample_91$rail),
    sd_rail <- mean(sample_91$rail)
  )



print(sample_stats)
```

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)

The outlier days, suggests times when the number of people riding wmata was significantly higher or lower than the average levels. Those values could indicate several things for example: holidays, severe weather, major events like a sports game, or transit line disruptions. I would examine those specific days and research what happened that day, why does the data look like this, is it a pattern and why does it matter?
---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
md_car_thetfs <- read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
view(md_car_thetfs)
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

ANSWER: a per capita rate is more useful because areas with large populations will inherently have more thefts. Calculating the rate by 10,000 people allows for a fair comparison.

```{r}
md_car_thetfs|> 
  mutate(theft_rate <- (thefts_2023 / population) * 1000)
  
```

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
median_rate <- md_car_thefts |>
  summarise(median_rate = median(theft_rate)) |>

total_thefts <- md_car_thetfs|>
  summarise(total_thefts = sum(thefts))
total_thefts <- md_car_thefts |>
  summarise(total_thefts = sum(thefts))

counties_above_median <- md_car_thefts |>
  filter(theft_rate > median_rate)

print(total_thefts)
print(counties_above_median)
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

---

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)

To investigate an editor's suspicion of underpredorming pools, I would probably want to prioritize calculating the median and the IQR. While the mean is helpful, the median gives a better sense of the middle value because it's less affected by extreme outliers. The IQR shows the difference between the third and first quartiles—measures the spread of the middle 50% of the data and helps identify potential outliers using the 1.5 times IQR rule. Finding these outliers provides the statistical evidence needed to determine if specific pools are actually failing compared to typical performance.

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)

I'd would first calculate a z-score for each pool by subtracting the mean of 187 from its attendance and dividing by the standard deviation of 42. This standardized score shows precisely how many standard deviations a pool is from the average, which reveals which ones are potentially outside 99.7% of expected data. Z-scores are way more useful than raw rankings because they place all groups on a standard scale allowing for a fair comparison. A raw ranking only tells you who's last, whereas a z-score tells you if that pool is just a normal variation or actually a crazy statistical anomaly.

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

I'd first ask the department to describe its methodology to make sure they used a simple random sample rather than a convenience sample, which could create sampling bias if they only tracked easy days. I'd also ask if there was undercoverage, specifically if any part of the summer season was excluded, since that would make the 187 average unrepresentative of the actual population. In the story, I'd probably raise concerns about sampling error, explaining that because these figures are based on 60 days rather than a full count, they represent a cone of uncertainty rather than absolute certainty.

---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
