---
title: "PollingSimulation.Rmd"
output: html_document
date: "2026-03-03"
---
title: "Polling Simulation: Sampling and the Central Limit Theorem"
---

In this activity, you'll use data from the 2024 Cooperative Election Study (CES) — a large-scale national survey conducted during the 2024 election — to simulate what happens when a news organization tries to estimate public opinion from a poll sample. The full dataset has nearly 55,000 respondents, which we'll treat as our population.

Your goal is to see firsthand how sample size affects the accuracy of estimates, and what the distribution of sample estimates actually looks like — the core idea behind the Central Limit Theorem.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from this URL and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

We'll focus on one question: what share of American adults identify as religiously unaffiliated (atheist, agnostic, or nothing in particular)? This has been a major story in American religion journalism for years.

First, we'll create a new column called `unaffiliated` that is `TRUE` if a respondent's `trad` value is `"Atheist"`, `"Agnostic"`, or `"Nothing in Particular"`, and `FALSE` otherwise. Then calculate the proportion of the full dataset that is unaffiliated. Save this value as `population_proportion`.

```{r}
ces <- ces |>
  mutate(unaffiliated = trad %in% c("Atheist", "Agnostic", "Nothing in Particular"))

proportion <- mean(ces$unaffiliated)
```

That population proportion is the "true" population value that any poll is trying to estimate. Write it down — you'll be comparing everything else to it.

### Task 4: Take a single small sample

Use `set.seed(42)` and then sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of unaffiliated respondents in your sample.

```{r}
mean(ces$unaffiliated)
set.seed(42)

sample_50 <- ces |> sample_n(50)

proportion_50 <- mean(sample_50$unaffiliated)

print(sample_50)
print(proportion_50)
```

How close is your sample estimate to the original proportion? Is it surprising?

ANSWER HERE 

My sample estimate was 0.32 which was .03 off of the original proportion, not all that surprising considering the size of the sample.

### Task 5: Take a single larger sample

Now sample 500 rows. Set a different seed value, grab the sample and save it as `sample_500`. Calculate the proportion of unaffiliated respondents in this sample.

```{r}
set.seed(58)

sample_500 <- ces |> sample_n(500)

proportion_500 <- mean(sample_500$unaffiliated)

print(proportion_500)


```

How did the estimate change compared to the sample of 50? What does this suggest about sample size?

ANSWER HERE

This time the estimate changed by 0.014 compared to the 0.32 number we got from the sample of 50, it aligns with what we already know: that broadening the sample size leads to a more reliable and consistent estimate.

### Task 6: Simulate many polls

A single sample can get lucky or unlucky. To understand how polling actually works, we need to see the *distribution* of estimates across many samples. The code below draws 1,000 samples of size 50, calculates the proportion of unaffiliated respondents in each, and stores the results.

```{r}
set.seed(123)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$unaffiliated)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)

print(simulations_50)
print(simulations_500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side — one for the 1,000 samples of size 50 and one for the 1,000 samples of size 500. Add a vertical line at `proportion` to each so you can see how well the samples are centering on the true value.

```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share Religiously Unaffiliated",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```

### Task 8: Reflect on what you see

**What shape do both histograms have?** Describe them in plain language. Does it surprise you that this shape emerges from a variable that is simply TRUE or FALSE for each respondent?

Both histograms are pretty much bell-shaped and centered around the true proportion. Its kind of surprising that simple True of False questions create such a a normal looking distribution. 

**Compare the two histograms.** What is the most visible difference between the n=50 and n=500 distributions? What does that difference tell you about what happens when a poll uses a larger sample?

The most visible difference between the graphs is that n=50's distributions is much more spread out, while n=500's is much narrower and clustered. This tells me that larger sample sizes produce more precise estimates.

**Both distributions are centered near the red line.** What does it mean that 1,000 different random samples — each one drawn independently — tend to cluster around the same value? What would it mean for journalism if that *weren't* the case?

Because samples tend to cluster around the red line, random sampling produces estimates close to the true value. If the that wasn't the case, polls would probably not a reliable resource/tool for journalists.

**A reporter tells you their story is based on a poll of 50 people.** Based on what you've seen in the simulation, what questions would you want to ask before trusting that number? You don't need to use any technical terms — just describe what you observed.


If I saw a reporter using a sample size that small I would begin by asking how the sampled was selected, questioning the randomness of the sample. I would go further to compare those numbers to other polls to see if similar results are being obtained before trusting the immediate conclusion.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
