---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Sol Garcia-Terrazas"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

ANSWER HERE

There aren't any obvious data issues like weird or impossible values, the data set seems pretty clean compared to the Pennsylvania one. But all the FARMS percentages are really high and close together (79%-96%), so there's not much variation. That is a problem because it makes it harder to see a clear relationship between FARMS and math scores.


### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

ANSWER HERE

The scatter plot shows a negative relationship, as the FARMS percentage goes up, the percent of students meeting math expectations goes down. The points follow a strong downward trend, although there's still some spread in the middle ranges. I think compared to the Pennsylvania plot, this relationship looks stronger and more consistent, with a a clear direction and less random scatter. Montgomery County, MD is slightly larger and more economically diverse, which could help explain the stronger relationship between FARMS and math scores, though other factors like data quality, school policies, or demographics could also play a role.

### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

ANSWER HERE

The p-value is basically 0 (statistically significant), the adjusted R-squared is 0.673, and the coefficient for FARMS pct is -0.595. This means there is a strong and statistically significant negative relationship between poverty and math scores, as FARMS increases math performance clearly decreases. The model explains about 67% of the variation in scores. This model is definitely worth using because it shows a strong, reliable pattern. Compared to the Pennsylvania model, this one is much stronger.

### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

ANSWER HERE
Largest positive residual: Fox Chapel Elementary (20876)
Fox Chapel Elementary is located in Germantown, MD. The area is fairly diverse and has a large population of lower and middle-income families. There is also a large Hispanic population which suggests a more immigrant-heavy community. The school is performing much better than expected, which suggests something inside the school, like strong teaching, leadership, or support programs, is helping students succeed. My first question to the principal would be: What specific strategies or programs are helping your students outperform expectations despite economic challenges?

Largest negative residual: Germantown Elementary (20874)
This schools is also in Germantown, MD which tends to be diverse and mixed-income. Based on the model, the school should be preforming much better than it actually is, meaning something beyond economic challenges/poverty is affecting outcomes. This could be due to differences in resources, school climate, or student support systems. My first reporting question would be: What challenges is your school facing that might explain why students are preforming below expectations compared to similar schools?

### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

ANSWER HERE

The Maryland data strongly supports the school board member's claims, showing a clear and statistically significant negative relationship between FARMS rates and math scores. Schools with higher poverty levels tend to perform worse. This pattern is much stronger than the one seen in the Pennsylvania model, where the relationship was inconsistent. If I called a school board member, I would ask whether they think poverty alone explains performance differences, or if school-level factors should also be considered. The most interesting finding is Fox Chapel Elementary, which significantly outperforms the model despite high FARMS rates, making it a strong candidate for a visit. I'd want to see what that school is doing differently to succeed.
