---
title: "Multiple Regression Analysis: Maryland Voter Turnout"
name: "Sol Garcia-Terrazas"
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships in civic data, such as how different demographic and socioeconomic factors collectively impact voter turnout. By understanding these relationships, journalists can provide deeper insights into civic participation patterns and identify potential barriers to voting.

```{r setup}
library(tidyverse)
#install.packages('corrplot')
install.packages("corrplot")
library(corrplot)
```

## Load the Data

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains information about voter turnout and various demographic factors across Maryland counties. This dataset will be used to build our model that will try to see which factors are most important in predicting voter turnout. The turnout percentage is from the 2024 general election.

```{r}
md_turnout <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_turnout.csv")
head(md_turnout)
```

## Task 1: Exploratory Data Analysis (5 points)

Before building our model, let's explore the data to understand the variables and their relationships. Examine the correlation matrix to see how the independent variables relate to each other and to the dependent variable (voter turnout). Which variables seem to be correlated? Are there any that are not correlated at all? Also look at the multiple scatterplots. Which variables seem to have a linear relationship with the dependent variable? Which do not?

ANSWER HERE

Looking at the correlation matrix, a few variables clearly stand out. Voter turnout has a strong negative relationship with percent unemployed (-0.74) and a strong positive relationship with the white pct (0.69). It's also moderatly related to median age (0.50) and income (0.45). Some of the independent variables like income and percent with a bachelor's degree (0.75) were also pretty correlated. Some variables are not related at all: like percent with a bachelor's degree and median age.

In the scatter plots, the pct unemployed and pct white have the clearest linear relationships with turnout. Income and median age also show an upward shift, but its not that strong. Population and pct with a bachelor's degree show up a lot more scattered, showing a weaker relationship with turnout. 



```{r}
# Summary statistics
summary(md_turnout)

# Create correlation matrix
selected_vars <- md_turnout |> 
  select(turnout, population, income, pct_bachelors_higher, 
         pct_unemployed, median_age, pct_white)
correlation_matrix <- cor(selected_vars)
print(correlation_matrix)

# Visualize correlation matrix
corrplot(correlation_matrix, method = "circle")

# Create scatterplots
md_turnout |>
  pivot_longer(cols = c(population, income, pct_bachelors_higher, 
                       pct_unemployed, median_age, pct_white),
               names_to = "variable", 
               values_to = "value") |>
  ggplot(aes(x = value, y = turnout)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Relationship between predictors and voter turnout",
       y = "Voter turnout (%)")
```

## Task 2: Initial Model Evaluation (5 points)

Next, we construct an initial linear model to analyze how different factors influence voter turnout across Maryland counties. This model helps us quantify the relationships and assess which factors are significant predictors of turnout. 

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For this model:

* **Min and Max** values indicate the range of residuals. 
* **Median** value tells us if the model is generally centered around the actual data points.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

* **Population**: Is the coefficient positive or negative? Is it statistically significant?
* **Income**: Does higher income correlate with higher turnout?
* **Education (pct_bachelors_higher)**: How does education level impact voter participation?
* **Unemployment**: Is there a relationship between unemployment and voter turnout?
* **Median Age**: Do counties with older populations have different turnout rates?
* **Race (pct_white)**: Is there a relationship between racial composition and turnout?

### Model Fit

* **Residual Standard Error**: How much does the model typically deviate from actual turnout?
* **Multiple R-squared**: What percentage of the variability in turnout is explained by the model?
* **Adjusted R-squared**: How well does the model fit after accounting for the number of variables?
* **F-statistic**: Is the model as a whole statistically significant?


Examine the output of the model, checking the coefficients, residuals, and overall model fit. How well does this multiple variable model predict turnout? Do any of the variables seem to be insignificant? Is this a useful model, based on what you've seen? Why?

ANSWER HERE

I think this model does a pretty good job of predicting voter turnout. The residuals are small and centered around zero, and the r-squared (0.85-0.80) shows the model explains most of the variation in turnout. The coefficients, income, median age, and pct white are significant and positively related to turnout. On the other hand variables like: population, pct with a bachelor's degree and pct unemployed, are not significant. While the model is useful, it could be probably simplified by removing less important variables.


```{r}
# Build initial model with all predictors
initial_model <- lm(turnout ~ population + income + pct_bachelors_higher + 
                   pct_unemployed + median_age + pct_white, data = md_turnout)
summary(initial_model)

```

## Task 3: Model Refinement (5 points)

Create three alternative models by adding or removing variables to see if we can improve the model's performance. Turnout will always be the dependent variable, but you should alter the independent variables by removing or adding some. Then compare the models using adjusted R-squared values. Are any of the models significantly better than the others?

ANSWER HERE

Model 2 focuses on the strongest predictors by keeping income, median age and pct white while removing weaker variables. Model 3 tests whether unemployment can explain turnout by including it instead of median age, and Model 4 replaces income with education to see if pct with a bachelor's degree preforms differently. 

After testing these models, none are dramatically better than the first. Model 2 has the highest r-squared value (0.8024), but it's the same as the original model. Model 3 performs slightly worse (0.766), and Model 4 drops off the most (0.657). Removing weaker variables didn't do as much as I expected it would, Model 2 is the best choice because of its simplicity but it works just as well as the original.


```{r}
# Model 2 (remove insignificant variables)
model2 <- lm(turnout ~ income + median_age + pct_white,  data = md_turnout)
summary(model2)

# Model 3 (keeping unemployment)
model3 <- lm(turnout ~ income + pct_unemployed + pct_white, data = md_turnout)
summary(model3)

# Model 4 (education instead of income)
model4 <- lm(turnout ~ pct_bachelors_higher + median_age + pct_white, data = md_turnout)
summary(model4)

# Compare models
cat("Model 1 (Full model) - Adjusted R-squared:", summary(initial_model)$adj.r.squared, "\n")
cat("Model 2 - Adjusted R-squared:", summary(model2)$adj.r.squared, "\n")
cat("Model 3 - Adjusted R-squared:", summary(model3)$adj.r.squared, "\n")
cat("Model 4 - Adjusted R-squared:", summary(model4)$adj.r.squared, "\n")

# Create comparison chart
model_names <- c("Full model", "Model 2", "Model 3", "Model 4")
adj_r_squared <- c(summary(initial_model)$adj.r.squared,
                   summary(model2)$adj.r.squared,
                   summary(model3)$adj.r.squared,
                   summary(model4)$adj.r.squared)

model_comparison <- data.frame(Model = model_names, Adjusted_R_squared = adj_r_squared)
ggplot(model_comparison, aes(x = reorder(Model, Adjusted_R_squared), y = Adjusted_R_squared)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison by Adjusted R-squared",
       x = "Model")
```


## Task 4: Using the Model (3 points)

Without using the words "multiple regression model", describe how you could use this analysis to cover an election or voter participation story.

ANSWER HERE

This analysis could help to explain why turnout higher in some counties than other by showing which factors matter the most like: income, age and demographics. For an election story, I could use it to compare counties with high and low turnout and investigate the patterns in those differences. This analysis help guide reporting by showing where to look for a story. 
