---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```



### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
m_ms <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
print(m_ms)
```



### Task 3: Sample 5 rows of data from the dataset

```{r}
set.seed(42)

sample_5 <- m_ms|> sample_n(5)

print(sample_5)
```



### Task 4: Calculate the mean age and standard deviation of the sample, using either the blue or white number of M&Ms

```{r}
sample_mean <- mean(sample_5$black)
sample_sd <- sd(sample_5$black)

sample_mean
sample_sd
```



### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4

```{r}
population_mean <- mean(m_ms$black)
population_sd <- sd(m_ms$black)

population_mean
population_sd
```



### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

The sample and population numbers are somewhat similar, but definitely skewed. They differ in that the sample mean (80.2) is higher than the population mean (73.3), while the sample sd (9.8) is significantly lower than the population sd (15.7).

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?
In the data sample stats were a bit off from the population stats. What really stood out was how the sample mean was 7 off and the standard deviation at least 5. Those inconsistencies show that small random samples do not always accurately reflect the full population data, they give reasonable estimates but the bigger the sample size the more accurate.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
