---
title: "HW8b: Polling Simulation"
---

In class, you simulated polling estimates for the share of Americans who are religiously unaffiliated — a group that makes up about 36% of the population. In this assignment, you'll run the same simulation for a different group: White Evangelicals, who make up roughly 10% of the population.

That difference matters more than it might seem. When a proportion is close to 50%, samples tend to be pretty well-behaved. When it's much smaller — or much larger — samples can behave differently, especially at small sizes. Your job is to run the simulation, compare what you find to the in-class results, and think carefully about what it means for reporting on religion and politics.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)

```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from the same URL as the in-class activity and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

Create a new column called `evangelical` that is `TRUE` if a respondent's `trad` value is `"White Evangelical"` and `FALSE` otherwise. Then calculate the proportion of the full dataset that is White Evangelical and save it as `population_proportion`.

```{r}
ces <- ces |> 
  mutate(evangelical = trad == "White Evangelical")
population_proportion <- mean(ces$evangelical)

population_proportion
```

Write down this value. How does it compare to the unaffiliated proportion (about 0.36) from the in-class activity?

ANSWER HERE

The population proportion of White Evangelicals in the data set is about 0.10, much lower than the unaffiliated proportion of around 0.36. This is probably because a smaller number of the survey respondents identify as White Evangelicals than unaffiliated.

### Task 4: Take a single small sample

Use `set.seed(42)` and sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of White Evangelical respondents in your sample.

```{r}
set.seed(42)

sample_50 <- ces|> sample_n(50)

sample_proportion <- mean(sample_50$evangelical)

sample_proportion
```

How close is your sample estimate to the population proportion? How does the accuracy compare to what you saw with the unaffiliated group in class?

ANSWER HERE

The sample estimate is around 0.16, higher than the population proportion of 0.10. This means that the sample size statistic overestimates the number of White Evangelical respondents in the full data set, this is most likely due to the fact that the sample size is so small. This is similar to what we saw in our in-class activity with the unaffiliated group, it ultimatly relates back to the Central Limit Theorem.

### Task 5: Take a single larger sample

Use `set.seed(1966)` and sample 500 rows. Save it as `sample_500`. Calculate the proportion of White Evangelical respondents in this sample.

```{r}
set.seed(1966)

sample_500 <- ces |> sample_n(500)

sample_propotion500 <- mean(sample_500$evangelical)

sample_propotion500
```

Did the larger sample improve the estimate? How does the improvement compare to what you saw in class?

ANSWER HERE
The larger sample size definitely improved the estimate, this time the sample proportion was 0.9 which is pretty close to the population proportion of 0.10. This shows a larger sample is best for more accurate results, which is exactly what we saw in the in-class activity with the disaffiliated group.

### Task 6: Simulate many polls

The code below runs the same simulation as the in-class activity, but using your `evangelical` column. Run it as-is.

```{r}
set.seed(42)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$evangelical)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)

simulations_50
simulations_500
```

### Task 7: Visualize the simulation results

Create two histograms side by side using the code structure from class. Add a vertical line at `population_proportion`. Make sure to update the title and subtitle to reflect that this simulation is about White Evangelicals.

```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = population_proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share White Evangelical",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```

### Task 8: Reflect on what you see

**Describe the shape of the n=50 histogram.** Is it as symmetric as the unaffiliated histogram from class? Why do you think that is, given that white evangelicals make up only about 10% of the population?

ANSWER HERE

The n=50 histogram is bell shaped but not perfectly symmetric, like the in class histogram we made of the unaffiliated responses the data is spaced out weirdly. It's skewed because white evangelicals make up only 10% of the data set. With small sample groups and the small proportions of the white evangelicals, random variation causes the estimates to be less balanced around the true value.  

**Compare the n=50 and n=500 histograms for this group.** How much does the distribution change with a larger sample? Is the improvement bigger or smaller than what you saw in class with the unaffiliated group?

ANSWER HERE

With a larger sample size of 500, the distribution of data is way more narrow and concentrated around the mean. It's a much cleaner and accurate way to read the data and it reflects a smaller improvement compared to the histograms we made in-class, probably due to the small proportion of white evangelicals in the larger data set.

**A news organization wants to report on White Evangelical voters using a poll of 200 people.** Based on your simulations, roughly how many White Evangelical respondents would that poll be likely to include? What does that suggest about the reliability of estimates for smaller subgroups?

ANSWER HERE

If white evangelicals are about 10% of the population, then a poll of 200 people would most likely be around 20 white evangelical respondents. The small number means that the subgroup is subject to a lot of variation from poll to poll. This shows how estimates for smaller groups like the white evangelicals are less reliable as they are more affected by random variation.

**Imagine you are covering the 2028 presidential election and a campaign shares a poll showing White Evangelical support at 78%.** The poll sampled 400 people. Based on everything you've seen in this assignment and in class, what would you want to know before reporting that number?

ANSWER HERE

Before reporting that number, I would want to know how many white evangelicals were included in the poll, if they make up 10% of the population, a sample of 400 people should have around 40 white evangelicals. I would also ask about the margin of error for that specific subgroup, and how that sample was collected. This is crucial because small subgroups like that are more prone to random variation and you don't want to report an incorrect number.


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
