---
title: "Final Project: College Spending and Graduation Rates"
by: "Sol Garcia-Terrazas" 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Ideas of Tests to Run
Data Prep
- Filtering out missing data for accurate conclusions
- Filter by institution type (2-year vs. 4-year), so sperate analysses can be done avoiding invalid comparisons.
Tests
- Correlation analysis to measure the strength of the relationship between spending per award and graduation rates
- Linear regression, to test is whether the relationship is statistically significant, im not very confident in this part.
- Comparing means of graduation rates (public vs. private institutions) : summary statistics and t-testing
- Potentially focusing on group means (for example racial groups) to investigate potential disparities.
Visualizations
- Scatter plot (spending award vs graduation rate): to identify patterns and outliers.
- Scatter plot with regression line?
- Boxplot: for graduation rates by institution type, public vs. private


## Introduction of Final Project

Updated Main Hypothesis: Colleges with higher spending per award tend to have higher graduation rates than those that spend less. However it might reflect differences in institutional resources rather than student investment. 

Additional Hypothesis: The relationship between spending per award and graduations rates differ between public and private institutions

For my final project I am using data from: https://www.kaggle.com/datasets/thedevastator/boost-student-success-with-college-completion-da/data

It's based on U.S Department of Education data covering variables like graduation rates, spending per student, tuition, enrollment, demographics, and admission data and other institutional features from the mid 2010s to the early 2020s.

I am focusing on the variables: 
- exp_award_value <- Spending per award
- grad_150_rate <- main outcome variable
- control <- public vs. private institutions
- level <- 2 year vs. 4 year institutions (does degree type affect the outcome, analyses conducted within levels)
- student_count: measures total enrollment (does institution size matter?)
- race and gender <- could potentially reveal disparities in graduation rates across different student groups.


potential ideas: 
- scatterplot (spending per award vs graduation rate)
- Compare means (public vs private)

```{r}
library(tidyverse)

inst_grads <- read_csv("data/cc_institution_grads.csv")
inst_details <- read_csv("data/cc_institution_details.csv")
state_grads <- read_csv("data/cc_state_sector_grads.csv")
state_details <- read_csv("data/cc_state_sector_details.csv")


glimpse(inst_grads)
glimpse(inst_details)
```



#Joining the Data sets were focusing on inst_details and state_details

```{r}
inst_full <- inst_grads |>
  left_join(inst_details, by = "unitid")

glimpse(inst_full)
summary(inst_full)

clean_data <- inst_full |>
  filter(!is.na(exp_award_value), !is.na(grad_150_rate))
```


# What am I missing? What are my limitations.

The data set does not include a direct measure of spending per student which was part of my original idea. Now I am using exp_award_value (spending per award) as a proxy, which may not perfectly capture overall institutional spending.

Also the data set lacks variables related to student income, financial aid or tuition, which could limit essential context.

How should I handle missing data?

