---
title: "Final Project: College Spending and Graduation Rates"
by: "Sol Garcia-Terrazas" 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction of Final Project

Updated Main Hypothesis: Colleges with higher spending per award tend to have higher graduation rates than those that spend less. However it might reflect differences in institutional resources rather than student investment. 

Additional Hypothesis: The relationship between spending per award and graduations rates differ between public and private institutions

For my final project I am using data from: https://www.kaggle.com/datasets/thedevastator/boost-student-success-with-college-completion-da/data

It's based on U.S Department of Education data covering variables like graduation rates, spending per student, tuition, enrollment, demographics, and admission data and other institutional features from the mid 2010s to the early 2020s.

I am focusing on the variables: 
- exp_award_value <- Spending per award
- grad_150_rate <- main outcome variable
- control <- public vs. private institutions
- level <- 2 year vs. 4 year institutions (does degree type affect the outcome, analyses conducted within levels)
- student_count: measures total enrollment (does institution size matter?)
- race and gender <- could potentially reveal disparities in graduation rates across different student groups.


potential ideas: 
- scatterplot (spending per award vs graduation rate)
- Compare means (public vs private)

## Ideas of Tests to Run
Data Prep
- Filtering out missing data for accurate conclusions
- Filter by institution type (2-year vs. 4-year), so sperate analysses can be done avoiding invalid comparisons.
Tests
- Correlation analysis to measure the strength of the relationship between spending per award and graduation rates
- Linear regression, to test is whether the relationship is statistically significant, im not very confident in this part.
- Comparing means of graduation rates (public vs. private institutions) : summary statistics and t-testing
- Potentially focusing on group means (for example racial groups) to investigate potential disparities.
Visualizations
- Scatter plot (spending award vs graduation rate): to identify patterns and outliers.
- Scatter plot with regression line?
- Boxplot: for graduation rates by institution type, public vs. private
Updated/Additional Analyses
- Simple linear regression:  
  - to clearly estimate the spending-per-award coefficient and see if they relate to grad rates.
- Multiple regression: 
  - Public vs. private (control), 2-year vs. 4-year (level), and student count (size). 
  - This is to test whether the relationship between spending and grad rates hold after controlling for institutional differences.
- Outlier identification:
  - Z-score used to identify:
    - High-preforming institutions (above expected graduation rates)
    - Low-preforming institutions (below expected graduation rates)
- Running separate models by group (public vs. private, 2-year vs. 4-year)


# What am I missing? What are my limitations.

The data set does not include a direct measure of spending per student which was part of my original idea. Now I am using exp_award_value (spending per award) as a proxy, which may not perfectly capture overall institutional spending.

Also the data set lacks variables related to student income, financial aid or tuition, which could limit essential context.

How should I handle missing data?

```{r}
library(tidyverse)

inst_grads <- read_csv("data/cc_institution_grads.csv")
inst_details <- read_csv("data/cc_institution_details.csv")
state_grads <- read_csv("data/cc_state_sector_grads.csv")
state_details <- read_csv("data/cc_state_sector_details.csv")


glimpse(inst_grads)
glimpse(inst_details)
```


# Joining the Data sets were focusing on inst_details and state_details

```{r}
inst_full <- inst_grads |>
  inner_join(inst_details, by = "unitid")

glimpse(inst_full)
summary(inst_full)

clean_data <- inst_full |>
  filter(
    !is.na(exp_award_value),
    !is.na(grad_150_rate),
    !is.na(control),
    !is.na(level),
    !is.na(student_count)
  )
clean_data <- clean_data |> mutate(spend_k = exp_award_value / 1000)
# reference groups
clean_data$level <- relevel(as.factor(clean_data$level), ref = "4-year")
```


# Filtering by institution type

```{r}
four_year <-clean_data |>
  filter(level == "4-year")

two_year <-clean_data |>
  filter(level == "2-year")

count (clean_data,level)
```


# Correlation Analysis

Correlation: 0.1244046, positive but very weak relationship 

```{r}
cor(clean_data$spend_k, clean_data$grad_150_rate, use = "complete.obs")
```

# Simple Linear Regression
Residuals: (this model has low explanatory power)
     Min       1Q   Median       3Q      Max 
-307.147  -24.756   -3.061   20.721   63.992 

Model Equation:
Graduation rate = 36.01 + 0.05133 x spending award <- (spend_k)

Coefficient:
spend_k = 0.05133 (for every $1,000 increase in spending per award -> graduation rates increase by 0.05133)

p-value = < 0.00000000000000022 <- statistically significant but not practically meaningful.

R-squared = 0.01548 <- thats pretty low and shows that spending only explains 1.5% of the variation in graduation rates

Conclusion from this test: spending more is associated with slightly higher graduation rates, the effect however is VERY SMALL.
```{r}
simple_model <- lm(grad_150_rate ~ spend_k, data = clean_data)
summary(simple_model)
```
# Multiple Regression
Intercept (Baseline: private for profit, 4-year private institution) = 47.15 <- 4-year schools have a predicted graduation rate of about 47%

spend_k = 0.02341 <- positive but smaller than before (0.05133), once we control for school type and size, spending matters even less.

Institution type
controlPublic = -18.29 <- public institutions 18 percentage points lower graduation rates than private for-profit institutions
controlPrivate not-for-profit = +0.76 <- slightly above the estimate but small effect.

Level:
level2-year = -5.03 <- 2-year schools have a 5% lower graduation rates than 4-year institutions

Size:
student_count = 0.0001654 <-larger school have SLIGHTLY higher grad rates; the effect is SMALL

R-squared = 0.1236 <- this model explains about 12.4% of variation in graduation rates (compared to the simple model of 1.5%, institutional characteristics matter way more than spending alone)

Reflection: Most variation is explained by other factors not in this model

```{r}
multiple_model <- lm(
  grad_150_rate ~ spend_k + control + level + student_count,
  data = clean_data
)
summary(multiple_model)
```

# Public Vs. Private graduation rates
mean_grad_rate:
Private for-profit: 46.20%
Private not-for-profit: 50.28% <- the advantage of private not-for profit institutions in raw avergages dissapears after controlling for institutional characteristics, I found that interesting.
Public: 28.81% <- public schools have much lower graduation rates than both types of private schools (17-21% lower)

median_grad_rate:
Private for-profit: 48.3
Private not-for-profit: 50
Public: 25 

```{r}
clean_data |>
  group_by(control) |>
  summarise(
    mean_grad_rate = mean(grad_150_rate),
    median_grad_rate = median(grad_150_rate),
  )

```

# T-test comparing average grad rate of 4-year schools vs. 2-year schools
Welch Two Sample t-test:

Means:
4-year schools = 44.05
2-year schools = 30.88 <- 13.2% difference

t = 179.31
p-value < 0.00000000000000022 <- statistically significant observed gap

95% confidence interval: 13.03 - 13.32 <- 95% sure that 4-year institutions have grad rates 13.0 to 13.3% higher than 2-year schools. (smaller difference than the regression model which said it was 5% higher) <- does that matter?

The larger difference found in the t-test compared to the regression model matters because the t-test compares 2-year and 4-year institutions without controlling for other factors. The regression model adjusts for variables like spend_k, enrollment size and institutional control, suggesting that part of the graduation-rate gap is explained by broader institutional differences

```{r}
t.test(grad_150_rate ~ level, data = clean_data)
```
# Outlier Identification Using Z-Scores

Okay the first time I ran this, I messed up and didn't filter out duplicate columns, so the data wasn't super accurate at first. What I decided to do was to aggregate the data to school level before identifying outliers, this was to prevent the same institution from appearing repeatedly.

I later counted the number of 2-year and 4-year schools to make sure the new table accurately reflected the original datasets proportions. This helped verify that the aggregated data set did not distrort the overall distribution of school types. (is this the right thing to do?)

RESULTS: 

Several schools have average graduation rates of 0%, including:
- Argosy University <- appears three times in the bottom 10
- Capella University
- Strayer University
- Rasmussen College-Wisconsin	
- Keweenaw Bay Ojibwa Community College	
* These are predominatly for-profit institutions, often with non-traditional or online student pops.

Some institutions report moderate to high spending per award, but still have extremeley low graduation rates, such as: 
- Keweenaw Bay Ojibwa Community College	($126,733 per award, 0% graduation rate)
- Argosy University Campuses ($60,000 - $90,000 per award, 0% graduation rate)
* Meaning higher spending per award does not guarantee better graduation outcomes 

There are a few other schools thar have slightly higher, although still very low graduation rates:
- Ohio Mid-Western College (1.17%, while spending $271,188 per award)
- University of Phoenix at Richmond (2.42%, while spending $156,574 per award)

Schools with extremely high graduation rates (above 90%), including:
- Harvard University (96.32%)
- Princeton University (95.22%)
- Yale University (94.52%)
- Amherst College	(94.18%)
- Pomona College (93.62)
* Predominantly elite private not-for-profit institutions, usually with highly selective admissions processes

Many of these top-preforming schools also have very high spending awards:
- Yale University ($556,675) spend_k <- 556.675	
- Princeton University ($383,472) spend_k <-383.472
- Harvard University ($381,077) spend_k <- 381.077
* Suggests that higher spending per award is associated with strong outcomes in some cases, especially among well-resourced institutions. 

Important Outlier:
- University of Management and Technology
  - Graduation Rate: 100%
  - Avg Spending per award: $2,059
  
In this data set:
- Top preforming schools have z-scores around +2.6 to +3.0, indicating rates far above the average
- Low performing schools have z-scores around -2.0, indicating rates far below the average

```{r}
school_level_data <- clean_data|>
  group_by(unitid, chronname, control,level)|>
  summarise(
    avg_grad_150_rate = mean(grad_150_rate, na.rm = TRUE),
    spend_k = mean(spend_k, na.rm = TRUE),
    student_count = mean(student_count, na.rm = TRUE),
    .groups = "drop"
  )

school_level_data <- school_level_data|>
  mutate(
    grad_zscore = as.numeric(scale(avg_grad_150_rate))
  )

school_level_data |> 
  arrange(desc(grad_zscore)) |>
  select (unitid,chronname,avg_grad_150_rate, spend_k, control, level, grad_zscore)|>
  head(10)

school_level_data |>
  arrange(grad_zscore)|>
  select(unitid,chronname,avg_grad_150_rate, spend_k
         , control, level, grad_zscore)|>
  head(10)

count(school_level_data, level)
count(clean_data, level)
```
# Separate Models by Group

Public Institution Model:

Baseline = 34.60 <- Public 4-year institutions have an estimated graduation rate of about 34.6%
exp_award_value = 0.000026974 <- has a positive effect, but its very small

Level: 2-year vs. 4-year
2-year estimate: -16.34 <- Public 2-year schools have about 16% lower graduation rates than their 4-year counterparts.
*This is larger than the multiple regression model (-5%), meaning this gap is stronger in public institutions

student_count = 0.000193840 <- larger public schools have slightly higher graduation rates, SMALL EFFECT

Rsquared = 0.1305 <- the model explains about 13% of the variations in grad rates (stronger than my simple model and similar to my multiple model)


```{r}
public_model <- clean_data|>
  filter(control == "Public") |>
  lm(grad_150_rate ~ exp_award_value + level + student_count, data = _)

summary(public_model)

```


#Private Institution Model: (Includes for-profit and not-for-profit)
Baseline: 44.11 <- a private for profit 4-year institution has an estimated graduation rates of 44%

exp_award_value = 0.000031823 <- positive but small effect, larger than the public model

level2-year = 11.13 <- private 2-year schools have 11% higher graduation rates than private 4-year schools.
*This is the opposite of what we saw in the public model (-16%). What could explain this is that Private 2-year schools are often career-focused, certificate programs or just generally have shorter completion time (making it easier to graduate within 150% time)

student_count = 0.000058333 <- larger private schools have slightly higher graduation rates, VERY SMALL effect

r-squared = 0.02321 <- this model explains 2.3% of the variation in private school graduation rates (MUCH smaller than the public model of 13%)

```{r}
private_model <- clean_data|>
  filter(control != "Public") |>
  lm(grad_150_rate ~ exp_award_value + level + student_count, data = _)

summary(private_model)
```


#4-year Institutions Model:

Baseline = 33.54% <- a 33.5% estimated grad rate for a private for-profit 4-year institution

exp_award_value = 0.0000426647 <- spending per award has a positive effect (slightly stronger than in my full multpile model: 0.000023)

Institution Type:
controlPrivate not-for-profit = 11.91 <- private not for-profit 4-year schools have 12% higher grad rates than private for-profit schools. *not-for-profit schools clearly outperform schools among 4-year institutions

controlPublic = -1.47 <- public 4-year schools have -1.47% lower graduation rates than private for-profit schools (MUCH smaller than the 18% gap from the full multiple model)
*the large gap from earlier is mostly about institution or degree type

student_count = 0.0003376207 <- larger 4-year schools have slightly higher grad rates, small effect

r-squared = 0.06381 <- this model explains 6% of variation in graduation rates 

```{r}
four_year_model <- clean_data|>
  filter(level == "4-year")|>
  lm(grad_150_rate ~ spend_k + control + student_count, data = _)
summary(four_year_model)

```


#2-year Institutions Model:

Baseline = 59.15% <- represents the predicted graduation rate for a private for-profit 2-year institution 

spend_k = -0.04407 <- for every additional $1,000 spent per award, graduation rates decrease by 0.04 percentage points (after controlling for institutional type and enrollment size) This is a sharp contrast from the other models.

controlPrivate not-for-profit = -7.865 <- private not-for-profit 2 year institutions have graduation rates  7.9 percentage points lower than private for-profit institutions. This suggests institutional type may influence graduation outcomes independently of spending levels. 

controlPublic = -33.91 <- public 2-year graduation rates about 34 percentage points lower than  private for-profit institutions. This is one of the strongest effects in the model and could reflect the larger, more economically diverse student populations often served by public community colleges.

student_count = -0.0003201 <- larger 2-year are associated with slightly lower graduation rates; THE PRACTICAL EFFECT IS VERY SMALL (enrollment size alone has little impact on graduation outcomes).

p-value < 0.001 <- all variables are statistically significant

r-squared = 0.3298 <- this model explaiuns about 33% of the variation in graduation rates among 2-year institutions (MUCH LARGER than the other models)

CONCLUSION: Unlike the overall and 4-year models, the 2-year model shows a small negative relationship between spending per award (spend_k) and graduation rates. This suggests that higher spending alone may not improve outcomes at 2-year institutions, where broader structural and socioeconomic factors may play a larger role in student success. 


```{r}
two_year_model <- clean_data|>
  filter(level == "2-year")|>
  lm(grad_150_rate ~ spend_k + control + student_count, data = _)
summary(two_year_model)

```



#Sampled scatterplot: spending vs. graduation rate

I first tried to create a regular scatterplot with a regression line. However because the dataset contains hundreds of thousands of observations, the visualization became heavily overplotted and very difficult to interpret. So I decided to pivot and make a sampled scatterplot for easier readability



OLD SCATTERPLOT -  Meant to show the error first made

At first the data was only showing/compressed in a very small corner, I had to find out how to fix this so I looked up the standard rstudio ggplot guide on ggplot2.tidyverse.org where I found the Cartesian coordinates function which zooms in the plot. When I was able to fix that error, I ran into another one. The data set was way too big. To make the visualization more understandable, I created a sampled scatterplot using a subset of the data.


source used for troubleshooting: 
https://ggplot2.tidyverse.org/reference/coord_cartesian.html
```{r}
ggplot(clean_data, aes(x = spend_k, y = grad_150_rate)) +
  geom_point(alpha = 0.05, size = 1) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  coord_cartesian(xlim = c(0, 500), ylim = c(0, 100)) +
  labs(
    title = "Spending per Award and Graduation Rates",
    x = "Spending per Award (in $1,000s)",
    y = "Graduation Rate"
  ) 
```


Sampling the Data: REVISED VISUALIZATION
```{r}
set.seed(123)

sample_data <- clean_data |>
  sample_n(5000)

sample_data |>
  ggplot(aes(x = spend_k, y = grad_150_rate)) +
  geom_point(alpha = 0.4, size = 1.2) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  coord_cartesian(xlim = c(0, 500) , ylim = c(0, 100)) +
  labs(
    title = "Relationship Between Spending per Award and Graduation Rates",
    x = "Spending per Award (in $1,000s)",
    y = "Graduation Rate"
  ) 
```
Findings: The scatter plot shows a slight positive relationship relationship between spending per award and graduation rates. As spending per award increases, graduation rates tend to increase sightly. Because the data is widely scattered, it suggest that while there might be a positive relationship between spending per award and graduation rates, spending alone is not a strong predictor of graduation outcomes

upward regression line: positive relationship

points are very spread out: weak relationship/low correlation



#Boxplot: graduation rates by institution control/type
```{r}
ggplot(clean_data, aes(x = control, y= grad_150_rate)) +
  geom_boxplot() +
  labs (
    title = "Graduation Rates by Institution Type",
    x = "Institution Type/Control",
    y = "Graduation Rate", 
  )

```
Findings: The box plot shows differences in graduation rates across institution types. private not-for-profit institutions have the highest median graduation rate, followed pretty closely by private for-profit institutions. Public institutions have the lowest median graduation rate, though several high-value outliers that indicate that there are several public institutions that are preforming exceptionally well.

# Final Section - Story Pitch

Story Pitch:
This story examines whether colleges that spend more per award tend to have higher graduation rates. This analysis foud a statistically significant but weak positive relationship between spending and graduation rates. This suggests that spending alone is not a strong predictor of graduation rates.

Further analysis shows that institutional characteristics tend to matter more. Public institutions and 2-year schools generally had lower graduation rates than private and 4-year institutions, its important to note that these differences became smaller after controlling for institutional factors like enrollment size and school type, suggesting that broader structural differences between schools may explain much of the variation.

Major outliers were also identified. Elite private universities like Harvard, Princeton and Yale combined extremely high graduation rates with very high spending per award. On the other end of the spectrum several institutions like Ohio Mid-Western College reported very low graduation rates despite relatively high spending levels, further reinforcing that spending does not guarantee strong student outcomes.


Potential Problems and Limitations:
Several limitations affect the analysis of this data. The dataset does not include direct measures of student income, financial aid access, academic preparation or the selectivity of an institution -- all of which likely influence graduation outcomes.

Spending per award is also only a proxy for institutional investment and may not fully capture how resources are being distributed to students.

The results focus on identifying statistical relationships. Because the analysis is observational, it doesn't determine whether spending causes higher graduation rates. 


Reporting Plan:
My story would focus on the gap between institutional spending and student outcomes. My reporting would compare high-preforming and low-preforming schools, examine differences between public and private institutions and explore why higher spending does not always translate into better student outcomes. 

My reporting would include: 
- The visuals I created in rstudio (Box plot and Scatter plot)
- Interviews with education researchers or administrators about why spending alone may not predict student success, and if not what does affect those outcomes?
- A comparison of structural inequalities and differences in student populations across institutions.
